---
title: "MSI-SRT"
author: "Ludvig Larsson"
date: '2022-06-16'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries

```{r message=FALSE, warning=FALSE}
library(STutility)
library(ggplot2)
library(magrittr)
library(dplyr)
library(dbscan)
```


Assemble spaceranger output files and MSI output files

# Create directories and system variables
These directories should have already been created if you already ran the main.Rmd script. Run the following chunk of 
code in case you didn't. Modify the project.dir and working.dir paths depending on the folder where you are located 
when running this script. You can check the location where you are using running getwd() directly in the console and 
modify it using setwd() function. When writing this script we are working in the /sma/scripts folder (our working.dir), 
and we counsidered sma our project.dir.

```{r message=FALSE, warning=FALSE}

project.dir <- ".."
data.dir <- paste0(project.dir, "/data")
working.dir <- "."
dir.create(paste0(working.dir, "/R_objects"))
dir.create(paste0(working.dir, "/results"))
dir.create(paste0(working.dir, "/results/tables"))
dir.create(paste0(working.dir, "/results/plots"))

```

Assemble spaceranger output files and msi csv files.

```{r}

# spaceranger output files
samples <- list.files(path = data.dir, pattern = "filtered_feature_bc_matrix.h5", recursive = TRUE, full.names = TRUE)
imgs <- list.files(path = data.dir, pattern = "tissue_hires_image.png", recursive = TRUE, full.names = TRUE)
spotfiles <- list.files(path = data.dir, pattern = "tissue_positions_list.csv", recursive = TRUE, full.names = TRUE)
json <- list.files(path = data.dir, pattern = "scalefactors_json.json", recursive = TRUE, full.names = TRUE)
metrics <- list.files(path = data.dir, pattern = "metrics_summary.csv", recursive = TRUE, full.names = TRUE)

# Collect sample IDs from file names for Visium datasets
sampleids <- do.call(rbind, strsplit(samples, "/"))
sampleids <- sampleids[, ncol(sampleids) - 4]

# Create infoTable for STUtility
infoTable <- data.frame(samples, imgs, spotfiles, json, sample_id = sampleids)

# MSI data files
msi_files <- list.files(path = data.dir, pattern = "_smamsi.csv", recursive = TRUE, full.names = TRUE)

# Collect sample IDs from file names for MSI datasets
sampleids_msi <- do.call(rbind, strsplit(msi_files, "/"))
sampleids_msi <- sampleids_msi[, ncol(sampleids_msi) - 3]

# Add ample IDS as names to msi files
msi_files <- setNames(msi_files, nm = sampleids_msi)

```

There is another issue that remains and that is how to align the MSI data with the Visium data for the human brain samples. Even though the strategy that we used is the same, we aligned this data in another script (MSI_SRT_hPDStr_LLMV.Rmd). Please refer to this script to see the parameters that we used for the alignment of the human data. For this script we will remove the human samples (V11T17-102 slide)

```{r}

msi_files <- msi_files[-grep(pattern = "V11T17-102", x = names(msi_files))]

```

We'll only keep sample ids with paired Visium and MSI data. 

```{r}

# Samples to keep
samples_keep <- intersect(sampleids, names(msi_files))

# Filter infoTable
infoTable_filtered <- subset(infoTable, sample_id %in% samples_keep)

# Filter msi files
msi_files_filtered <- msi_files[samples_keep]

```

## Load MSI data

Here we also name the list based on sample IDS so that it's 
easier to pair it the Visium data with the MSI data later on. 
This step should take no more than a minute to run.

```{r}

# Create mepty lists
msiList <- list()
coordList <- list()

# Fill lists with coordinates and count matrices
for (nm in  names(msi_files_filtered)) {
  f <- msi_files_filtered[nm]
  msiData <- data.table::fread(input = f, sep = ",")
  xy <- as.data.frame(msiData[, 1:2]) # Extract coordinates
  msiData <- msiData[, 3:ncol(msiData)] # Remove coordinates from msiData
  msiData <- t(msiData) # Transpose data
  colnames(msiData) <- paste0(xy$x, "x", xy$y) # Set x,y coords as rownames
  rownames(xy) <- paste0(xy$x, "x", xy$y)
  msiList <- c(msiList, setNames(list(msiData), nm = nm))
  coordList <- c(coordList, setNames(list(xy), nm = nm))
}

```

Each element in `msiList` contains an MSI matrix and each element in 
`coordList` contains a `data.frame` with x, y coordinates.

```{r}

msiList[[1]][1:5, 1:5]
head(coordList[[1]])

```

Check that lists match

```{r}

# Check that names match
sprintf("Do all element names of the lists match? %s", all(names(msiList) == names(coordList)))

```

## Create RNA Seurat object
***

Here we'll create a Seurat object for each Visium dataset. It'll take 
a couple of minutes to load all datasets.

```{r eval=FALSE}

# Load Visium data
se.RNA.list <- setNames(lapply(1:nrow(infoTable_filtered), function(i) {
  se <- InputFromTable(infoTable_filtered[i, ])
  return(se)
}), nm = infoTable_filtered$sample_id)

```

The default image size for an H&E image (tissue_hires_image.png) is roughly 2000x2000 pixels. 
So in order to make use of STUtility's alignment function we need to adjust the MSI coordinates
to be similar. To achieve this, I decided to multiply the MSI grid coordinates by a factor of 25.
We will see later what effect this transformation has.


```{r}

sf <- 20

coordList <- lapply(coordList, function(coords) {
  coords <- cbind(coords, setNames(coords*sf, nm = c("pixel_x", "pixel_y")))
  return(coords)
})

```

Now let's have a look at the coordinates for Visium and MSI in the same plot 
and color these by the number of unique features:

```{r fig.width=10, fig.height=20}

plots <- lapply(names(se.RNA.list), function(nm) {
  
  se.RNA <- se.RNA.list[[nm]]
  coords <- coordList[[unique(se.RNA$sample_id)]]
  msiMat <- msiList[[unique(se.RNA$sample_id)]]
  gg <- rbind(cbind(GetStaffli(se.RNA)@meta.data[, c("pixel_x", "pixel_y")], 
                    type = "RNA", unique_features = se.RNA$nFeature_RNA),
           cbind(coordList[[unique(se.RNA$sample_id)]][, c("pixel_x", "pixel_y")], 
                 type = "MSI", unique_features = colSums(msiMat > 0)))

  plots <- lapply(c("RNA", "MSI"), function(i) {
    inner_p <- ggplot(subset(gg, type == i), aes(pixel_x, pixel_y, color = unique_features)) +
      geom_point() +
      scale_color_gradientn(colours = viridis::viridis(n = 11)) +
      theme_void() +
      theme(legend.position = "top", legend.text = element_text(angle = 60, hjust = 1)) +
      labs(title = paste0(nm, ": ", i))
    return(inner_p)
  })
  
  p <- patchwork::wrap_plots(plots, ncol = 2)
  
  return(p)
  
})

patchwork::wrap_plots(plots, ncol = 2)

```

We can also visualize this as violin plots:

```{r fig.height=7, fig.width=10}

gg <- do.call(rbind, lapply(names(se.RNA.list), function(nm) {
  se.RNA <- se.RNA.list[[nm]]
  coords <- coordList[[unique(se.RNA$sample_id)]]
  msiMat <- msiList[[unique(se.RNA$sample_id)]]
  gg <- rbind(cbind(GetStaffli(se.RNA)@meta.data[, c("pixel_x", "pixel_y")], 
                    type = "RNA", unique_features = se.RNA$nFeature_RNA),
           cbind(coordList[[unique(se.RNA$sample_id)]][, c("pixel_x", "pixel_y")], 
                 type = "MSI", unique_features = colSums(msiMat > 0)))
  gg$sample_id <- nm
  return(gg)
}))

p1 <- ggplot(subset(gg, type == "MSI"), aes(sample_id, unique_features, fill = sample_id)) +
  geom_violin(scale = "width") +
  geom_hline(yintercept = 200, linetype = "dashed", color = "red") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
p2 <- ggplot(subset(gg, type == "RNA"), aes(sample_id, unique_features, fill = sample_id)) +
  geom_violin(scale = "width") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

patchwork::wrap_plots(p1, p2, ncol = 1) &
  theme(legend.position = "none")

```

## Create Seurat objects for MSI data
***

Now we can create a `Staffli` object and a `Seurat` object for each dataset to and combine the two to make a 
`Seurat` object compatible with STUtility. 

```{r}

# Create Seurat objects for MSI data
se.MSI.list <- setNames(lapply(names(se.RNA.list), function(nm) {
  coords <- coordList[[nm]]
  xy_lims <- sapply(coords[, 1:2], max)
  # Create a Staffli object and set sample to 1 since we only have 1 tissue section
  st.object <- CreateStaffliObject(meta.data = cbind(coords, sample = "1"))
  # Do limits vary between experiments?
  st.object@limits[["1"]] <- xy_lims
  # Create Seurat object
  se.MSI <- CreateSeuratObject(counts = msiList[[nm]], assay = "MSI")
  se.MSI@tools$Staffli <- st.object
  return(se.MSI)
}), nm = names(se.RNA.list))

```

It looks like the MSI section has some data outside of the tissue. We will get rid of it using a strategy that 
involves looking at the first components of the PCA.

```{r fig.width=10, fig.height=5}

(ST.FeaturePlot(se.RNA.list[[1]], features = "nFeature_RNA", palette = "viridis", pt.size = 2) | 
  ST.FeaturePlot(se.MSI.list[[1]], features = "nCount_MSI", palette = "viridis", pt.size = 2)) & theme(strip.text = element_blank())

```

```{r include=FALSE}

# Only used to speed up knitting process
se.MSI.list <- readRDS("../R_objects/se.MSI.list")

```

Let's subset the MSI data to have more than 250 features per pixel.

Then we run a few simple steps to process the data:

1. normalization 
2. scaling
3. detect top variable features
4. PCA

```{r eval=FALSE}

se.MSI.list <- lapply(se.MSI.list, function(se.MSI) {
  se.MSI <- SubsetSTData(se.MSI, expression = nFeature_MSI > 250)
  se.MSI <- se.MSI |> 
    NormalizeData(verbose = FALSE) |>
    ScaleData(verbose = FALSE) |>
    FindVariableFeatures(verbose = FALSE) |>
    RunPCA(verbose = FALSE)
  return(se.MSI)
})

```

If we look at the PCA, it seems as if the first components drives a separation 
between tissue and non-tissue. It makes sense that PC1 also shows up in the 
ventricles which should be empty of cells. 

```{r fig.width=9, fig.height=12}

selected.compontents <- setNames(c(1, 1, 2, 1, 1, 1, 1, 1, 1, 1, 1, 1), nm = names(se.MSI.list))

plots <- lapply(names(se.MSI.list), function(nm) {
  se.MSI <- se.MSI.list[[nm]]
  ST.DimPlot(se.MSI, dims = selected.compontents[nm], reduction = "pca", ncol = 2, 
             cols = RColorBrewer::brewer.pal(n = 11, name = "RdBu") |> rev())
})

patchwork::wrap_plots(plots, ncol = 3)

```

Based on these selected components, we can filter our MSI dataset 
further to get a better outline of the tissue. 

All thresholds were selected by visual inspection of the spatial plots above. 

```{r}

# The integer determines the PC and the sign
# determines the direction that the edge effect is in
selected.pcs.threshold <- setNames(c(40, 40, -25, 20, 20, 20, -40, 40, 40, 40, -40, -40), nm = names(se.MSI.list))

gg <- do.call(rbind, lapply(names(se.MSI.list), function(nm) {
  se.MSI <- se.MSI.list[[nm]]
  npc <- selected.compontents[nm]
  dir <- sign(selected.pcs.threshold[nm])
  thr <- selected.pcs.threshold[nm]
  if (dir > 0) {
    df <- data.frame(value = se.MSI@reductions$pca@cell.embeddings[, npc], sample_id = nm) |>
      mutate(tissue = ifelse(value > thr, "background", "tissue"))
  } else {
    df <- data.frame(value = se.MSI@reductions$pca@cell.embeddings[, npc], sample_id = nm) |>
      mutate(tissue = ifelse(value < thr, "background", "tissue"))
  }
  return(df)
}))

ggplot(gg, aes(sample_id, value, color = tissue)) +
  geom_jitter(size = 0.5) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

```

And let's look at the labels in 2D space:

```{r  fig.width=12, fig.height=12}

se.MSI.list <- setNames(lapply(names(se.MSI.list), function(nm) {
  se.MSI <- se.MSI.list[[nm]]
  npc <- selected.compontents[nm]
  dir <- sign(selected.pcs.threshold[nm])
  thr <- selected.pcs.threshold[nm]
  values <- se.MSI@reductions$pca@cell.embeddings[, npc]
  if (dir > 0) {
    se.MSI$tissue <- ifelse(values > thr, "background", "tissue")
  } else {
    se.MSI$tissue <- ifelse(values < thr, "background", "tissue")
  }
  return(se.MSI)
}), nm = names(se.MSI.list))

plots <- lapply(names(se.MSI.list), function(nm) {
  se.MSI <- se.MSI.list[[nm]]
  ST.FeaturePlot(se.MSI, features = "tissue", cols = c("background" = "black", "tissue" = "mistyrose")) &
    labs(title = nm) &
    theme(strip.text = element_blank())
})

patchwork::wrap_plots(plots, ncol = 3)

```

### Filter data based on PCA
***

Looks like our filters captured the edges quite well so now we can remove 
the background from our `Seurat` objects.

```{r}

se.MSI.list <- setNames(lapply(se.MSI.list, function(se.MSI) {
  se.MSI <- SubsetSTData(se.MSI, expression = tissue == "tissue")
  return(se.MSI)
}), nm = names(se.MSI.list))

```


## Align MSI data to SRT data
***

Unfortunately, `STUtility` is designed to work with image data so we cannot use 
the `Align*` functions directly on this object. To circumvent this issue, we could 
put in some "dummy" images into the `Staffli` object. To achieve this, we need to 
add a couple of things to our object.

First, let's load the H%E images for the Visium data.

```{r eval=FALSE}

se.RNA.list <- lapply(se.RNA.list, function(se.RNA) {
  se.RNA <- LoadImages(se.RNA, xdim = 400, time.resolve = FALSE)
  se.RNA <- se.RNA |> MaskImages(channels.use = 3)
  se.RNA <- SwitchResolution(se.RNA, xdim = 1000, verbose = TRUE)
  return(se.RNA)
})

```

Now the H&E images should be masked which is required to run 
manual image alignment later on. Let's have a look at the masked 
H&E images:

```{r fig.width=5, fig.height=5}

for (i in seq_along(se.RNA.list)) {
  GetStaffli(se.RNA.list[[i]])@rasterlists$masked[[1]] |> 
    magick::image_read() |>
    magick::image_scale(geometry = "400") |>
    as.raster() |>
    plot()
}

```


```{r fig.width=8, fig.height=10}

plots <- lapply(names(se.MSI.list), function(nm) {
  se.MSI <- se.MSI.list[[nm]]
  p <- ST.FeaturePlot(se.MSI, features = "nFeature_MSI", label.by = ) &
    theme(legend.position = "top", strip.text = element_blank()) &
    labs(title = nm)
  return(p)
})

patchwork::wrap_plots(plots, ncol = 3)

```

Now if we move to the MSI data instead, we need to add a few things to the
object to make it work in `STUtility`. 

```{r eval=FALSE}

se.MSI.withIMG.list <- setNames(lapply(names(se.MSI.list), function(nm) {
  
  # Make sure that image masks are removed
  se.MSI <- se.MSI.list[[nm]]
  se.RNA <- se.RNA.list[[nm]]
  st_rna_object <- GetStaffli(se.RNA)
  dim_row <- (st_rna_object@rasterlists$raw$`1` |> dim())[1]
  dim_col <- (st_rna_object@rasterlists$raw$`1` |> dim())[2]
  se.MSI@tools$Staffli@rasterlists$masked <- NULL
  se.MSI@tools$Staffli@rasterlists$masked.masks <- NULL
  
  # plot MSI coordinates to create a "tissue" image
  tmpfile <- tempfile()
  png(filename = paste0(tmpfile, ".png"), height = dim_row, width = dim_col)
  par(mar = c(0, 0, 0, 0))
  #p <- ST.FeaturePlot(se.MSI.filtered, features = "orig.ident", pt.size = 20, cols = "black") &
  #  theme(plot.title = element_blank(), plot.subtitle = element_blank(), legend.position = "none") 
  p <- cbind(se.MSI[[]], GetStaffli(se.MSI)@meta.data) |>
    mutate(pixel_x = pixel_x*0.5, pixel_y = pixel_y*0.5) |>
    ggplot(aes(pixel_x, dim_row - pixel_y)) +
    geom_point(size = 10) +
    scale_x_continuous(limits = c(0, dim_col), expand = c(0, 0)) +
    scale_y_continuous(limits = c(0, dim_row), expand = c(0, 0)) +
    theme_void()
  print(p)
  dev.off()
  
  # read "tissue" image
  im <- magick::image_read(path = paste0(tmpfile, ".png"))
  iminfo <- magick::image_info(im)
  im <- as.raster(im)
  
  # Now we can place the "tissue" image in our Staffli object
  # and we are ready to run manual image alignment
  se.MSI@tools$Staffli@rasterlists[["raw"]] <- list("1" = im)
  se.MSI@tools$Staffli@rasterlists[["masked"]] <- list("1" = im)
  se.MSI@tools$Staffli@rasterlists[["masked.masks"]] <- list("1" = as.raster(ifelse(im == "#ffffffff", "#00000000", "#ffffffff")))
  se.MSI@tools$Staffli@meta.data <- se.MSI@tools$Staffli@meta.data |> 
    mutate(pixel_x = pixel_x*0.5, pixel_y = pixel_y*0.5)
  se.MSI@tools$Staffli@xdim <- dim_col
  se.MSI@tools$Staffli@dims <- list("1" = iminfo)
  
  return(se.MSI)
  
}), nm = names(se.MSI.list))

```

As a sanity check, we can plot the MSI coordinates on our new "tissue" image
to see if they align:

```{r fig.width=12, fig.height=12}

plots <- lapply(names(se.MSI.list), function(nm) {
  FeatureOverlay(se.MSI.withIMG.list[[nm]], features = "nFeature_MSI", show.sb = FALSE, pt.size = 1)
})

patchwork::wrap_plots(plots, ncol = 3)
  
```


```{r eval=FALSE, include=FALSE}

saveRDS(se.MSI.withIMG.filtered, file = "../R_objects/se.MSI.withIMG.list")

```

Now we can merge the two objects `se.RNA` and `se.MSI.filtered` into one single
Seurat object. To do this, we use the `MergeSTData` function

```{r eval=FALSE}

se.merged.list <- lapply(names(se.MSI.withIMG.list), function(nm) {
  se.merged <- MergeSTData(se.RNA.list[[nm]], se.MSI.withIMG.list[[nm]])
  return(se.merged)
})

```

The two coordinate systems are fairly close to each other now.

```{r include=FALSE}
se.merged.list <- readRDS("../R_objects/se.merged.list")
```

Let's plot our paired RNA and MSI data points in a spatial map.

```{r fig.width=14, fig.height=14}

plots <- lapply(se.merged.list, function(se.merged) {
  se.merged$type <- ifelse(is.na(se.merged$nCount_MSI), "RNA", "MSI")
  p <- ST.FeaturePlot(se.merged, features = "orig.ident", ncol = 2, cols = "black", 
                      pt.size = 1, label.by = "type")
  return(p)
})

patchwork::wrap_plots(plots, ncol = 2)

```

### Manual alignment
***

With this we should be able to run `ManualAlignImages`. Here, we need to 
adjust the MSI section so that it fits with the RNA data section. This 
will be the most time consuming step of this notebook.

To align the data, we need to apply rotation, translations and distortions
using the shiny app opened by `ManualAlignImages`.

NOTE: The tissue outline for samples "V11T17-105_A1" and "V11T17-105_B1" 
are difficult to align, since parts of the tissue are cropped out. The 
alignment of these datasets might therefore be unreliable.

Here are the settings I used:
    
```{r}

transforms <- data.frame(
  sample_id = names(se.MSI.withIMG.list),
  rotation_angle = c(89.6, 91.2, 89.3, 89.1, 89.8, 89.1, 89.5, 89.8, 89.2, 89.4, 89, 89.6),
  move_along_x_axis = c(151, 163, 147, 161, 152, 215, 108, 1, 52, 186, 106, 66),
  move_along_y_axis = c(-175, -165, -139, -193, -100, -126, -96, -125, 6, -46, -154, -7),
  blue_axis_angle = c(50.8, 48.7, -49.2, 55.3, 55.3, 54.6, 53.4, 50, 48.7, 55.8, 52.3, 55.3),
  blue_axis_stretch = c(1.13, 1.12, 1.19, 1.18, 1.17, 1.17, 1.15, 1.17, 1.18, 1.18, 1.19, 1.19),
  red_axis_angle = c(49, 48.7, -49.3, -52.3, 55.2, 54.6, 53.4, 46.5, 45.6, 54.4, 42.3, 53.8),
  red_axis_stretch = c(1.19, 1.19, 1.2, 1.2, 1.22, 1.21, 1.19, 1.17, 1.19, 1.19, 1.2, 1.19)
)
transforms

```

Run the alignment function for each dataset.

```{r eval=FALSE}

se.merged.list[[1]] <- ManualAlignImages(se.merged.list[[1]])
se.merged.list[[2]] <- ManualAlignImages(se.merged.list[[2]])
se.merged.list[[3]] <- ManualAlignImages(se.merged.list[[3]])
se.merged.list[[4]] <- ManualAlignImages(se.merged.list[[4]])
se.merged.list[[5]] <- ManualAlignImages(se.merged.list[[5]])
se.merged.list[[6]] <- ManualAlignImages(se.merged.list[[6]])
se.merged.list[[7]] <- ManualAlignImages(se.merged.list[[7]])
se.merged.list[[8]] <- ManualAlignImages(se.merged.list[[8]])
se.merged.list[[9]] <- ManualAlignImages(se.merged.list[[9]])
se.merged.list[[10]] <- ManualAlignImages(se.merged.list[[10]])
se.merged.list[[11]] <- ManualAlignImages(se.merged.list[[11]])
se.merged.list[[12]] <- ManualAlignImages(se.merged.list[[12]])

```

Baed on the plots below, the alignment appears to be quite decent.

```{r fig.width=10, fig.height=10}

se.merged.list <- setNames(se.merged.list, nm = names(se.RNA.list))
plots <- lapply(names(se.merged.list), function(nm) {
  se.merged <- se.merged.list[[nm]]
  se.merged$sample_type <- ifelse(GetStaffli(se.merged)@meta.data$sample == "1", "RNA", "MSI")
  p <- ST.FeaturePlot(se.merged, features = "sample_type", ncol = 2, pt.size = 1.5) &
    theme(legend.position = "none", plot.background = element_rect(fill = "lightgray")) &
    labs(title = nm)
  return(p)
})

patchwork::wrap_plots(plots, ncol = 3)

```

Now that the images are aligned, we still have one last step left to put 
the MSI and Visium coordinates in the same "space".

From now on, we'll have the coordinates stored as a `data.frame` instead of 
using the `Seurat` objects. Below, we calculate a scaling factor to adjust 
the MSI coordinates to fit into the same coordinate system as the RNA data. 

```{r fig.width=12, fig.height=14}

aligned_coordinates <- do.call(rbind, lapply(names(se.merged.list), function(nm) {
  dims <- GetStaffli(se.merged.list[[nm]])@dims
  sf <- dims$`1`$width/dims$`2`$width
  coords <- cbind(se.merged.list[[nm]][[]], GetStaffli(se.merged.list[[nm]])@meta.data) |>
    tibble::rownames_to_column(var = "barcode") |>
    mutate(barcode = gsub(replacement = "_1", x = barcode, pattern = "_\\d")) |>
    select(barcode, warped_x, warped_y, nFeature_MSI) |>
    mutate(sample_type = case_when(is.na(nFeature_MSI) ~ "RNA",
                                   !is.na(nFeature_MSI) ~ "MSI")) |>
    mutate(warped_x = case_when(sample_type == "MSI" ~ warped_x*sf, TRUE ~ warped_x),
           warped_y = case_when(sample_type == "MSI" ~ warped_y*sf, TRUE ~ warped_y),
           sample_id = nm) |>
    select(barcode, warped_x, warped_y, sample_id, sample_type)
  return(coords)
}))

p <- ggplot(aligned_coordinates, aes(warped_x, warped_y,color = sample_type)) +
  geom_point(size = 0.5) +
  facet_wrap(~sample_id, ncol = 3)

p

```

```{r eval=FALSE}

# Export plot
p <- ggrastr::rasterize(p, layers = "Point", dpi = 300, dev = "ragg")
pdf("results/plots/aligned_coordinates_spatial_plot.pdf", width = 12, height = 14)
print(p)
dev.off()

```

## data integration
***

Now we need a strategy to pair data points across the two modalities. 

The strategy used downstream works as follows: For the MSI data pixels, assign at least 1 nearest neighbor 
from the RNA data spots and obtain the average expression for each gene

We also need to apply some kind of distance threshold to remove spots that are too far from a neighbor in 
the other data set.

First, let's calculate the pairwise distances between the two data sets:

```{r include=FALSE}

knn_spatial_df_list <- readRDS(file = "../R_objects/knn_spatial_df_list")

```


```{r eval=FALSE}

# Split coordinates into a list
aligned_coordinates_list <- split(aligned_coordinates, aligned_coordinates$sample_id)

# Find 5 closest spatial neighbors 
knn_spatial_df_list <- lapply(aligned_coordinates_list, function(coords) {
  
  # get spatial coordinates for RNA and MSI data
  set1 = subset(coords, sample_type == "RNA")[, c("barcode", "warped_x", "warped_y")] |>
    data.frame(row.names = 1)
  set2 = subset(coords, sample_type == "MSI")[, c("barcode", "warped_x", "warped_y")] |>
    data.frame(row.names = 1)
  
  # Calculate pairwise distances using k nearest neighbors algorithm
  knn_spatial_df <- kNN(x = set1, k = 5, query = set2)
  
  # Create a data.frame with MSI pixel IDs (from) pointing to its closest
  # Visium spot (to) and the distance between the two (distance)
  knn_spatial_df <- data.frame(from = rownames(set2)[rep(1:nrow(knn_spatial_df$id), 5)],
                                 to = rownames(set1)[as.vector(knn_spatial_df$id)],
                                 distance = as.vector(knn_spatial_df$dist))
  
  # Add spatial coordinates for "from" and "to"
  knn_spatial_df <- knn_spatial_df |>
    mutate(x = set2[from, "warped_x"],
           x_end = set1[to, "warped_x"],
           y = set2[from, "warped_y"],
           y_end = set1[to, "warped_y"])
  
  return(knn_spatial_df)
})

```

What we see in the plot below are the MSI data points in red 
and their closest RNA data neighbors in blue. We need to get 
rid of the MSI data points where we don't have any RNA data points.

```{r fig.width=12, fig.height=30}

plots <- lapply(names(knn_spatial_df_list), function(nm) {
  knn_spatial_df <- knn_spatial_df_list[[nm]]
  p <- ggplot() +
    geom_segment(data = knn_spatial_df, aes(x = x, xend = x_end, y = y, yend = y_end)) +
    geom_point(data = knn_spatial_df, aes(x, y), color = "red", size = 0.5) +
    geom_point(data = knn_spatial_df, aes(x_end, y_end), color = "blue", size = 0.5) +
    scale_y_reverse() +
    labs(title = nm)
  return(p)
})

p <- patchwork::wrap_plots(plots, ncol = 2)
p

```

```{r eval=FALSE, include=FALSE}

# Export plot
# this plot gets too big with PDF format because of all of the segments
for (i in seq_along(plots)) {
  jpeg(paste0("results/plots/nearest_neighbor_plots/", names(se.MSI.list)[i], ".jpeg"), width = 2000, height = 2000, res = 300)
  print(plots[[i]])
  dev.off()
}

```


If we look at the distance histogram, it should be relatively easy 
to spot what data points are located too far away from each other. 
A value of 35 looks like a reasonable cutoff based on the 
distance histogram below. 

```{r}

gg <- do.call(rbind, lapply(names(knn_spatial_df_list), function(nm) {
  knn_spatial_df <- knn_spatial_df_list[[nm]]
  knn_spatial_df$sample_id <- nm
  return(knn_spatial_df)
}))

ggplot(gg, aes(distance)) +
  geom_histogram(bins = 100) +
  scale_x_log10() +
  geom_vline(xintercept = 35, linetype = "dashed", color = "red")

```

Let's filter out data points that do not have a close match in the other data set:

```{r fig.width=5, fig.height=5}

knn_spatial_df_filtered_list <- lapply(knn_spatial_df_list, function(knn_spatial_df) {
  knn_spatial_df <- knn_spatial_df |>
    dplyr::filter(distance < 35)
  return(knn_spatial_df)
})

```

```{r fig.width=12, fig.height=30}

plots <- lapply(names(knn_spatial_df_filtered_list), function(nm) {
  knn_spatial_df <- knn_spatial_df_filtered_list[[nm]]
  p <- ggplot() +
    geom_segment(data = knn_spatial_df, aes(x = x, xend = x_end, y = y, yend = y_end)) +
    geom_point(data = knn_spatial_df, aes(x, y), color = "red", size = 0.5) +
    geom_point(data = knn_spatial_df, aes(x_end, y_end), color = "blue", size = 0.5) +
    scale_y_reverse() +
    labs(title = nm)
  return(p)
})

p <- patchwork::wrap_plots(plots, ncol = 2)
p

```

```{r eval=FALSE, include=FALSE}

# Export plot
# this plot gets too big with PDF format because of all of the segments
for (i in seq_along(plots)) {
  jpeg(paste0("results/plots/nearest_neighbor_filtered_plots/", names(knn_spatial_df_filtered_list)[i], ".jpeg"), width = 2000, height = 2000, res = 300)
  print(plots[[i]])
  dev.off()
}

```

First we filter the data to keep only 1 RNA data neighbor per MSI data 
point. This will lead to some loss of data, and is therefore a quite conservative approach.
An alternative would be to average the data between nearest neighbors, thus
interpolating values between data points.

```{r}

knn_spatial_df_filtered_list <- setNames(lapply(names(knn_spatial_df_filtered_list), function(nm) {
  
  knn_spatial_df_filtered <- knn_spatial_df_filtered_list[[nm]]
  
  res <- knn_spatial_df_filtered |> 
    dplyr::arrange(from, distance) |>
    dplyr::group_by(from) |>
    dplyr::slice_head(n = 1)
  
  sprintf("Spots removed from sample %s: %s", nm, sum(duplicated(res$to)))
  
  # Remove duplicated Visium spots 
  res <- res[!duplicated(res$to), ]
  
  return(res)
  
}), nm = names(knn_spatial_df_list))

```

Now, we can see that every data point in the MSI data has a match in the RNA data. 
However, with this current approach you can see that the regular pattern of the two
data types lead to a loss of data in certain parts of the tissue. 

```{r fig.width=12, fig.height=30}

plots <- lapply(names(knn_spatial_df_filtered_list), function(nm) {
  knn_spatial_df <- knn_spatial_df_filtered_list[[nm]]
  p <- ggplot() +
    geom_segment(data = knn_spatial_df, aes(x = x, xend = x_end, y = y, yend = y_end)) +
    geom_point(data = knn_spatial_df, aes(x, y), color = "red", size = 0.5) +
    geom_point(data = knn_spatial_df, aes(x_end, y_end), color = "blue", size = 0.5) +
    scale_y_reverse() +
    labs(title = nm)
  return(p)
})

p <- patchwork::wrap_plots(plots, ncol = 2)
p

```

```{r eval=FALSE, include=FALSE}

# Export plot
# this plot gets too big with PDF format because of all of the segments
for (i in seq_along(plots)) {
  jpeg(paste0("results/plots/nearest_neighbor_1RNA_1MSI_plots/", names(knn_spatial_df_filtered_list)[i], ".jpeg"), width = 2000, height = 2000, res = 300)
  print(plots[[i]])
  dev.off()
}

```

### Wrapping up the alignment process
***

Now that we have our aligned RNA and MSI coordinates, let's export these
as tables:

```{r}

for (nm in names(knn_spatial_df_filtered_list)) {
  write.csv(x = knn_spatial_df_filtered_list[[nm]], file = paste0("results/tables/", nm, "_nearest_neighbors.csv"), quote = FALSE, row.names = FALSE)
}

```

## Create multimodal assay
***

Now we can create a new `Seurat` object where we can analyze 
the two modalities jointly. Here we will simply add the MSI 
data to our RNA `Seurat` object.

```{r include=FALSE}

se.multi.list <- readRDS("../R_objects/se.multi.list")

```

Create a new `Seurat` object which holds both data modalities.

```{r eval=FALSE}

se.multi.list <- setNames(lapply(names(se.merged.list), function(nm) {
  knn_spatial_df_filtered <- knn_spatial_df_filtered_list[[nm]]
  se.multi <- se.RNA.list[[nm]]
  se.multi <- SubsetSTData(se.multi, spots = knn_spatial_df_filtered$to)
  MSI <- se.MSI.list[[nm]]@assays$MSI@counts[, knn_spatial_df_filtered$from]
  colnames(MSI) <- knn_spatial_df_filtered$to
  se.multi[["MSI"]] <- CreateAssayObject(counts = MSI)
  return(se.multi)
}), nm = names(se.merged.list))

```

Now we can also recalculate some basic QC metrics.

```{r eval=FALSE}

se.multi.list <- lapply(se.multi.list, function(se.multi) {
  se.multi$nFeature_RNA <- colSums(GetAssayData(se.multi, slot = "counts", assay = "RNA") > 0)
  se.multi$nFeature_MSI <- colSums(GetAssayData(se.multi, slot = "counts", assay = "MSI") > 0)
  return(se.multi)
})

```

And let's have a look at these QC metrics spatially.

```{r fig.width=6, fig.height=3}

ST.FeaturePlot(se.multi, features = c("nFeature_RNA", "nFeature_MSI"))

```

## Joint analysis
***

NOTE: I selected "CLR" normalization for the MSI data, but another 
normalization method might be more appropriate.

Below, we apply a basic workflow to normalize and analyze the RNA and MSI data sets. The strategy is outlined below:

1. Normalization ("logNormalize" RNA counts, "CLR" for "MSI" values)
2. Find top variable features
3. Run Principal Component Analysis for dimensionality reduction
4. Embed data with UMAP using the PCA results as input (first 30 components for the RNA data and first 10 components for the MSI data)
5. Create an SNN graph from the PCA results (first 30 components for the RNA data and first 10 components for the MSI data)
6. Run data-.driven clustering using default parameters

```{r eval=FALSE}

se.multi.list <- lapply(names(se.multi.list), function(nm) {
  
  se.multi <- se.multi.list[[nm]]
  cat(sprintf("Running analysis for sample %s", nm))
  
  # Normalize RNA data
  DefaultAssay(se.multi) <- 'RNA'
  se.multi <- NormalizeData(se.multi, verbose = FALSE) |> 
    FindVariableFeatures(verbose = FALSE) |> 
    ScaleData(verbose = FALSE) |> 
    RunPCA(verbose = FALSE) |>
    RunUMAP(reduction = "pca", dims = 1:30, reduction.name = "RNA.umap", verbose = FALSE) |>
    FindNeighbors(reduction = "pca", dims = 1:30, verbose = FALSE) |>
    FindClusters(verbose = FALSE)
  se.multi$RNA_clusters <- se.multi$seurat_clusters
  
  # Normalize MSI data
  DefaultAssay(se.multi) <- 'MSI'
  se.multi <- NormalizeData(se.multi, normalization.method = "CLR", verbose = FALSE) |> 
    FindVariableFeatures(nfeatures = 200, verbose = FALSE) |> 
    ScaleData(verbose = FALSE) |> 
    RunPCA(reduction.name = "msipca", verbose = FALSE) |>
    RunUMAP(reduction = "msipca", dims = 1:10, reduction.name = "MSI.umap", verbose = FALSE) |>
    FindNeighbors(reduction = "msipca", dims = 1:10, verbose = FALSE) |>
    FindClusters(verbose = FALSE)
  se.multi$MSI_clusters <- se.multi$seurat_clusters
  
  return(se.multi)
})

```

We can look at the UMAPs defined by either RNA or MSI data. It is clear that the RNA
data is more complex although it is still possible to find clustered data points in the
MSI data as well.

```{r fig.width=8, fig.height=30}

plots <- lapply(se.multi.list, function(se.multi) {
  p <- DimPlot(se.multi, group.by = "RNA_clusters", reduction = "RNA.umap") | 
    DimPlot(se.multi, group.by = "MSI_clusters", reduction = "MSI.umap")
  return(p)
})

patchwork::wrap_plots(plots, ncol = 1)

```

### Joint clustering
***

see [vignette](https://satijalab.org/seurat/articles/weighted_nearest_neighbor_analysis.html) for more details.

For each cell, we calculate its closest neighbors in the dataset based on a 
weighted combination of RNA and protein similarities. The cell-specific modality 
weights and multimodal neighbors are calculated in a single function, which takes 
~2 minutes to run on this data set. We specify the dimensionality of each modality 
(similar to specifying the number of PCs to include in clustering, but you can 
vary these settings to see that small changes have minimal effect on the overall 
results.

```{r}

# Identify multimodal neighbors. These will be stored in the neighbors slot, 
# and can be accessed using se.multi[['weighted.nn']]
# The WNN graph can be accessed at se.multi[["wknn"]], 
# and the SNN graph used for clustering at se.multi[["wsnn"]]
# Cell-specific modality weights can be accessed at se.multi$RNA.weight

se.multi.list <- lapply(se.multi.list, function(se.multi) {
  se.multi <- FindMultiModalNeighbors(
    se.multi, reduction.list = list("pca", "msipca"), 
    dims.list = list(1:30, 1:10), modality.weight.name = "RNA.weight"
  )
  return(se.multi)
})

```

We can now use these results for downstream analysis, such as visualization and 
clustering. For example, we can create a UMAP visualization of the data based on 
a weighted combination of RNA and protein data We can also perform graph-based 
clustering and visualize these results on the UMAP, alongside a set of cell annotations.

```{r eval=FALSE}

se.multi.list <- lapply(se.multi.list, function(se.multi) {
  se.multi <- RunUMAP(se.multi, nn.name = "weighted.nn", reduction.name = "wnn.umap", reduction.key = "wnnUMAP_", n.epochs = 1e3, min.dist = 0.15)
  se.multi <- FindClusters(se.multi, graph.name = "wsnn", algorithm = 3, resolution = 2, verbose = FALSE)
  se.multi$MM_clusters <- se.multi$seurat_clusters
  return(se.multi)
})

```

Now we can also sort the clusters by decreasing RNA weight.

```{r}

plots <- lapply(se.multi.list, function(se.multi) {
  p <- VlnPlot(se.multi, features = "RNA.weight", sort = TRUE)
  return(p)
})

patchwork::wrap_plots(plots[[3]], ncol = 1)

```

And we can compare the results for RNA only, MSI only and RNA+MSI side by side

```{r fig.width=12, fig.height=4}

plots <- lapply(se.multi.list, function(se.multi) {
  p1 <- DimPlot(se.multi, reduction = "RNA.umap", group.by = "RNA_clusters")
  p2 <- DimPlot(se.multi, reduction = "MSI.umap", group.by = "MSI_clusters")
  p3 <- DimPlot(se.multi, reduction = "wnn.umap", group.by = "MM_clusters")
  p <- p1 + p2 + p3 + patchwork::plot_layout(ncol = 3)
  return(p)
})

```

```{r fig.width=6, fig.height=6}

DefaultAssay(se.multi.list[[1]])
FeatureOverlay(se.multi.list[[5]], features = "MM_clusters")

```

```{r eval=FALSE, include=FALSE}

saveRDS(se.MSI.list, file = "R_objects/se.MSI.list")
saveRDS(se.RNA.list, file = "R_objects/se.RNA.list")
saveRDS(se.merged.list, file = "R_objects/se.merged.list")
saveRDS(se.multi.list, file = "R_objects/se.multi.list")
saveRDS(knn_spatial_df_list, "R_objects/knn_spatial_df_list")
saveRDS(knn_spatial_df_filtered_list, "R_objects/knn_spatial_df_filtered_list")

```


# Date
***

```{r}
date()
```

# Session
***

```{r}
devtools::session_info()
```

