---
title: "MSI-SRT_humanStriatum"
author: "Ludvig Larsson, Marco Vicari"
date: '2022-10-17'
output: html_document
---

This script uses the same strategy proposed in MSI_SRT_mPD_LL.Rmd to align the human Striatum MSI and SRT data. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries

```{r message=FALSE, warning=FALSE}
library(STutility)
library(ggplot2)
library(magrittr)
library(dplyr)
library(dbscan)
```


Assemble spaceranger output files and MSI output files

# Create directories and system variables
```{r message=FALSE, warning=FALSE}

project.dir <- "/home/marco.vicari/projects/sma_final_sub_230729/sma"
data.dir <- paste0(project.dir, "/data")
dir.create(paste0(project.dir, "/results"))
dir.create(paste0(project.dir, "/results/plots"))
plots.dir <- paste0(project.dir, "/results/plots")
dir.create(paste0(project.dir, "/results/tables"))
tables.dir <- paste0(project.dir, "/results/tables")
dir.create(paste0(project.dir, "/R_objects"))

```

Assemble spaceranger output files and msi csv files.

```{r}

# spaceranger output files
samples <- list.files(path = data.dir, pattern = "filtered_feature_bc_matrix.h5", recursive = TRUE, full.names = TRUE)
imgs <- list.files(path = data.dir, pattern = "tissue_hires_image.png", recursive = TRUE, full.names = TRUE)
spotfiles <- list.files(path = data.dir, pattern = "tissue_positions_list.csv", recursive = TRUE, full.names = TRUE)
json <- list.files(path = data.dir, pattern = "scalefactors_json.json", recursive = TRUE, full.names = TRUE)
metrics <- list.files(path = data.dir, pattern = "metrics_summary.csv", recursive = TRUE, full.names = TRUE)

# Collect sample IDs from file names for Visium datasets
sampleids <- do.call(rbind, strsplit(samples, "/"))
sampleids <- sampleids[, ncol(sampleids) - 4]

# Create infoTable for STUtility
infoTable <- data.frame(samples, imgs, spotfiles, json, sample_id = sampleids)

# MSI data files
msi_files <- list.files(path = data.dir, pattern = "_smamsi.csv", recursive = TRUE, full.names = TRUE)

# Collect sample IDs from file names for MSI datasets
sampleids_msi <- do.call(rbind, strsplit(msi_files, "/"))
sampleids_msi <- sampleids_msi[, ncol(sampleids_msi) - 3]

# Add ample IDS as names to msi files
msi_files <- setNames(msi_files, nm = sampleids_msi)

```

Here we select just the human samples (V11T17-102 slide)
```{r}

msi_files <- msi_files[grep(pattern = "V11T17-102", x = names(msi_files))]

```

We'll only keep sample ids with paired Visium and MSI data. 

```{r}

# Samples to keep
samples_keep <- intersect(sampleids, names(msi_files))

# Filter infoTable
infoTable_filtered <- subset(infoTable, sample_id %in% samples_keep)

# Filter msi files
msi_files_filtered <- msi_files[samples_keep]

```


## Load MSI data

Here we also name the list based on sample IDS so that it's easier to pair it the Visium data with the MSI data later on. 
This step should take no more than a minute to run.

```{r}

# Create mepty lists
msiList <- list()
coordList <- list()

# Fill lists with coordinates and count matrices
for (nm in  names(msi_files_filtered)) {
  f <- msi_files_filtered[nm]
  msiData <- data.table::fread(input = f, sep = ",")
  xy <- as.data.frame(msiData[, 1:2]) # Extract coordinates
  msiData <- msiData[, 3:ncol(msiData)] # Remove coordinates from msiData
  msiData <- t(msiData) # Transpose data
  colnames(msiData) <- paste0(xy$x, "x", xy$y) # Set x,y coords as rownames
  rownames(xy) <- paste0(xy$x, "x", xy$y)
  msiList <- c(msiList, setNames(list(msiData), nm = nm))
  coordList <- c(coordList, setNames(list(xy), nm = nm))
}

```

Each element in `msiList` contains an MSI matrix and each element in `coordList` contains a `data.frame` with x, y coordinates.

```{r}

msiList[[1]][1:5, 1:5]
head(coordList[[1]])

```

Check that lists match

```{r}

# Check that names match
sprintf("Do all element names of the lists match? %s", all(names(msiList) == names(coordList)))

```

## Create RNA Seurat object
***

Here we'll create a Seurat object for each Visium dataset. It'll take a couple of minutes to load all datasets.

```{r eval=FALSE}

# Load Visium data
se.RNA.list <- setNames(lapply(1:nrow(infoTable_filtered), function(i) {
  se <- InputFromTable(infoTable_filtered[i, ])
  return(se)
}), nm = infoTable_filtered$sample_id)

```

The default image size for an H&E image (tissue_hires_image.png) is roughly 2000x2000 pixels. So in order to make use of STUtility's alignment function we need to adjust the MSI coordinates to be similar. To achieve this, I decided to multiply the MSI grid coordinates by a factor of 25. We will see later what effect this transformation has.


```{r}

sf <- 25

coordList <- lapply(coordList, function(coords) {
  coords <- cbind(coords, setNames(coords*sf, nm = c("pixel_x", "pixel_y")))
  return(coords)
})

```

Now let's have a look at the coordinates for Visium and MSI in the same plot and color these by the number of unique features:

```{r fig.width=10, fig.height=20}

plots <- lapply(names(se.RNA.list), function(nm) {
  
  se.RNA <- se.RNA.list[[nm]]
  coords <- coordList[[unique(se.RNA$sample_id)]]
  msiMat <- msiList[[unique(se.RNA$sample_id)]]
  gg <- rbind(cbind(GetStaffli(se.RNA)@meta.data[, c("pixel_x", "pixel_y")], 
                    type = "RNA", unique_features = se.RNA$nFeature_RNA),
           cbind(coordList[[unique(se.RNA$sample_id)]][, c("pixel_x", "pixel_y")], 
                 type = "MSI", unique_features = colSums(msiMat > 0)))

  plots <- lapply(c("RNA", "MSI"), function(i) {
    inner_p <- ggplot(subset(gg, type == i), aes(pixel_x, pixel_y, color = unique_features)) +
      geom_point() +
      scale_color_gradientn(colours = viridis::viridis(n = 11)) +
      theme_void() +
      theme(legend.position = "top", legend.text = element_text(angle = 60, hjust = 1)) +
      labs(title = paste0(nm, ": ", i))
    return(inner_p)
  })
  
  p <- patchwork::wrap_plots(plots, ncol = 2)
  
  return(p)
  
})

patchwork::wrap_plots(plots, ncol = 2)

```

We can also visualize this as violin plots:

```{r fig.height=7, fig.width=10}

gg <- do.call(rbind, lapply(names(se.RNA.list), function(nm) {
  se.RNA <- se.RNA.list[[nm]]
  coords <- coordList[[unique(se.RNA$sample_id)]]
  msiMat <- msiList[[unique(se.RNA$sample_id)]]
  gg <- rbind(cbind(GetStaffli(se.RNA)@meta.data[, c("pixel_x", "pixel_y")], 
                    type = "RNA", unique_features = se.RNA$nFeature_RNA),
           cbind(coordList[[unique(se.RNA$sample_id)]][, c("pixel_x", "pixel_y")], 
                 type = "MSI", unique_features = colSums(msiMat > 0)))
  gg$sample_id <- nm
  return(gg)
}))

p1 <- ggplot(subset(gg, type == "MSI"), aes(sample_id, unique_features, fill = sample_id)) +
  geom_violin(scale = "width") +
  geom_hline(yintercept = 200, linetype = "dashed", color = "red") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
p2 <- ggplot(subset(gg, type == "RNA"), aes(sample_id, unique_features, fill = sample_id)) +
  geom_violin(scale = "width") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

patchwork::wrap_plots(p1, p2, ncol = 1) &
  theme(legend.position = "none")

```

Looking at these violin plots, it seems reasonable to think a bit more about QC steps. I doubt that the most sparse datasets would be useful for downstream 
analysis so it might be better to exclude some.  For now, let's keep all datasets for the data alignment.

## Create Seurat objects for MSI data
***
Now we can create a `Staffli` object and a `Seurat` object for each dataset to and combine the two to make a `Seurat` object compatible with STUtility. 

```{r}

# Create Seurat objects for MSI data
se.MSI.list <- setNames(lapply(names(se.RNA.list), function(nm) {
  coords <- coordList[[nm]]
  xy_lims <- sapply(coords[, 1:2], max)
  # Create a Staffli object and set sample to 1 since we only have 1 tissue section
  st.object <- CreateStaffliObject(meta.data = cbind(coords, sample = "1"))
  # Do limits vary between experiments?
  st.object@limits[["1"]] <- xy_lims
  # Create Seurat object
  se.MSI <- CreateSeuratObject(counts = msiList[[nm]], assay = "MSI")
  se.MSI@tools$Staffli <- st.object
  return(se.MSI)
}), nm = names(se.RNA.list))

```


```{r fig.width=10, fig.height=5}

(ST.FeaturePlot(se.RNA.list[[4]], features = "nFeature_RNA", palette = "viridis", pt.size = 2) | 
  ST.FeaturePlot(se.MSI.list[[4]], features = "nCount_MSI", palette = "viridis", pt.size = 2)) & theme(strip.text = element_blank())

```

```{r include=FALSE}

# # Only used to speed up knitting process
# se.MSI.list <- readRDS("../R_objects/se.MSI.list")

```

Let's subset the MSI data to have more than 250 features per pixel.

Then we run a few simple steps to process the data:

1. normalization
2. scaling
3. detect top variable features
4. PCA

```{r eval=FALSE}

se.MSI.list <- lapply(se.MSI.list, function(se.MSI) {
  se.MSI <- SubsetSTData(se.MSI, expression = nFeature_MSI > 250)
  se.MSI <- se.MSI |> 
    NormalizeData(verbose = FALSE) |>
    ScaleData(verbose = FALSE) |>
    FindVariableFeatures(verbose = FALSE) |>
    RunPCA(verbose = FALSE)
  return(se.MSI)
})

```

If we look at the PCA, it seems as if the first components drives a separation between tissue and non-tissue. It makes sense that PC1 also shows up in the 
ventricles which should be empty of cells. 

```{r fig.width=8, fig.height=2}

selected.compontents <- setNames(c(1, 1, 1,1), nm = names(se.MSI.list))

plots <- lapply(names(se.MSI.list), function(nm) {
  se.MSI <- se.MSI.list[[nm]]
  ST.DimPlot(se.MSI, dims = selected.compontents[nm], reduction = "pca", ncol = 2, 
             cols = RColorBrewer::brewer.pal(n = 11, name = "RdBu") |> rev())
})

patchwork::wrap_plots(plots, ncol = 4)

```

Based on these selected components, I think we can filter our MSI dataset further to get a better outline of the tissue. 
All thresholds were selected by visual inspection of the spatial plots above. 

```{r}

# The integer determines the PC and the sign
# determines the direction that the edge effect is in
selected.pcs.threshold <- setNames(c(-50, -40, -35, 30), nm = names(se.MSI.list))

gg <- do.call(rbind, lapply(names(se.MSI.list), function(nm) {
  se.MSI <- se.MSI.list[[nm]]
  npc <- selected.compontents[nm]
  dir <- sign(selected.pcs.threshold[nm])
  thr <- selected.pcs.threshold[nm]
  if (dir > 0) {
    df <- data.frame(value = se.MSI@reductions$pca@cell.embeddings[, npc], sample_id = nm) |>
      mutate(tissue = ifelse(value > thr, "background", "tissue"))
  } else {
    df <- data.frame(value = se.MSI@reductions$pca@cell.embeddings[, npc], sample_id = nm) |>
      mutate(tissue = ifelse(value < thr, "background", "tissue"))
  }
  return(df)
}))

ggplot(gg, aes(sample_id, value, color = tissue)) +
  geom_jitter(size = 0.5) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

```

And let's look at the labels in in 2D space:

```{r  fig.width=8, fig.height=2}

se.MSI.list <- setNames(lapply(names(se.MSI.list), function(nm) {
  se.MSI <- se.MSI.list[[nm]]
  npc <- selected.compontents[nm]
  dir <- sign(selected.pcs.threshold[nm])
  thr <- selected.pcs.threshold[nm]
  values <- se.MSI@reductions$pca@cell.embeddings[, npc]
  if (dir > 0) {
    se.MSI$tissue <- ifelse(values > thr, "background", "tissue")
  } else {
    se.MSI$tissue <- ifelse(values < thr, "background", "tissue")
  }
  return(se.MSI)
}), nm = names(se.MSI.list))

plots <- lapply(names(se.MSI.list), function(nm) {
  se.MSI <- se.MSI.list[[nm]]
  ST.FeaturePlot(se.MSI, features = "tissue", cols = c("background" = "black", "tissue" = "mistyrose")) &
    labs(title = nm) &
    theme(strip.text = element_blank())
})

patchwork::wrap_plots(plots, ncol = 4)

```

### Filter data based on PCA
***

Looks like our filters captured the edges quite well so now we can remove the background from our Seurat objects.

```{r}

se.MSI.list <- setNames(lapply(se.MSI.list, function(se.MSI) {
  se.MSI <- SubsetSTData(se.MSI, expression = tissue == "tissue")
  return(se.MSI)
}), nm = names(se.MSI.list))

```



## Align MSI data to SRT data
***

Unfortunately, `STUtility` is designed to work with image data so we cannot use the `Align*` functions directly on this object. To circumvent this issue, we could 
put in some "dummy" images into the `Staffli` object. To achieve this, we need to add a couple of things to our object.

First, let's load the H%E images for the Visium data.

```{r eval=FALSE}

se.RNA.list <- lapply(se.RNA.list, function(se.RNA) {
  se.RNA <- LoadImages(se.RNA, xdim = 400, time.resolve = FALSE)
  se.RNA <- se.RNA |> MaskImages(channels.use = 2)
  se.RNA <- SwitchResolution(se.RNA, xdim = 1000, verbose = TRUE)
  return(se.RNA)
})

```

Now the H&E images should be masked which is required to run manual image alignment later on. Let's have a look at the masked 
H&E images:

```{r fig.width=10, fig.height=5}

se.RNA.list <- lapply(se.RNA.list, function(se.RNA){
  msk <- se.RNA@tools$Staffli@rasterlists$masked$`1`
  msk[1, ] <- "#000000"
  msk[nrow(msk), ] <- "#000000"
  msk[, 1] <- "#000000"
  msk[, ncol(msk)] <- "#000000"
  se.RNA@tools$Staffli@rasterlists$masked$`1` <- msk
  msk <- se.RNA@tools$Staffli@rasterlists$masked.masks$`1`
  msk[1, ] <- "#000000"
  msk[nrow(msk), ] <- "#000000"
  msk[, 1] <- "#000000"
  msk[, ncol(msk)] <- "#000000"
  se.RNA@tools$Staffli@rasterlists$masked.masks$`1` <- msk
  return(se.RNA)
})


for (i in seq_along(se.RNA.list)) {
  par(mfrow = c(1, 3), mar = c(0,0,0,0))
  GetStaffli(se.RNA.list[[i]])@rasterlists$masked[[1]] |> 
    magick::image_read() |>
    magick::image_scale(geometry = "400") |>
    as.raster() |>
    plot()
  GetStaffli(se.RNA.list[[i]])@rasterlists$masked.masks[[1]] |> 
    magick::image_read() |>
    magick::image_scale(geometry = "400") |>
    as.raster() |>
    plot()
  GetStaffli(se.RNA.list[[i]])@rasterlists$raw[[1]] |> 
    magick::image_read() |>
    magick::image_scale(geometry = "400") |>
    as.raster() |>
    plot()
}

```


```{r fig.width=8, fig.height=2}

plots <- lapply(names(se.MSI.list), function(nm) {
  se.MSI <- se.MSI.list[[nm]]
  p <- ST.FeaturePlot(se.MSI, features = "nFeature_MSI", label.by = ) &
    theme(legend.position = "top", strip.text = element_blank()) &
    labs(title = nm)
  return(p)
})

patchwork::wrap_plots(plots, ncol = 4)

```

Now if we move to the MSI data instead, we need to add a few things to the
object to make it work in STUtility. 

```{r eval=FALSE}

se.MSI.withIMG.list <- setNames(lapply(names(se.MSI.list), function(nm) {
  
  # Make sure that image masks are removed
  se.MSI <- se.MSI.list[[nm]]
  se.RNA <- se.RNA.list[[nm]]
  st_rna_object <- GetStaffli(se.RNA)
  dim_row <- (st_rna_object@rasterlists$raw$`1` |> dim())[1]
  dim_col <- (st_rna_object@rasterlists$raw$`1` |> dim())[2]
  se.MSI@tools$Staffli@rasterlists$masked <- NULL
  se.MSI@tools$Staffli@rasterlists$masked.masks <- NULL
  
  # plot MSI coordinates to create a "tissue" image
  tmpfile <- tempfile()
  png(filename = paste0(tmpfile, ".png"), height = dim_row, width = dim_col)
  par(mar = c(0, 0, 0, 0))
  #p <- ST.FeaturePlot(se.MSI.filtered, features = "orig.ident", pt.size = 20, cols = "black") &
  #  theme(plot.title = element_blank(), plot.subtitle = element_blank(), legend.position = "none") 
  p <- cbind(se.MSI[[]], GetStaffli(se.MSI)@meta.data) |>
    mutate(pixel_x = pixel_x*0.5, pixel_y = pixel_y*0.5) |>
    ggplot(aes(pixel_x, dim_row - pixel_y)) +
    geom_point(size = 10) +
    scale_x_continuous(limits = c(0, dim_col), expand = c(0, 0)) +
    scale_y_continuous(limits = c(0, dim_row), expand = c(0, 0)) +
    theme_void()
  print(p)
  dev.off()
  
  # read "tissue" image
  im <- magick::image_read(path = paste0(tmpfile, ".png"))
  iminfo <- magick::image_info(im)
  im <- as.raster(im)
  
  # Now we can place the "tissue" image in our Staffli object
  # and we are ready to run manual image alignment
  se.MSI@tools$Staffli@rasterlists[["raw"]] <- list("1" = im)
  se.MSI@tools$Staffli@rasterlists[["masked"]] <- list("1" = im)
  se.MSI@tools$Staffli@rasterlists[["masked.masks"]] <- list("1" = as.raster(ifelse(im == "#ffffffff", "#00000000", "#ffffffff")))
  se.MSI@tools$Staffli@meta.data <- se.MSI@tools$Staffli@meta.data |> 
    mutate(pixel_x = pixel_x*0.5, pixel_y = pixel_y*0.5)
  se.MSI@tools$Staffli@xdim <- dim_col
  se.MSI@tools$Staffli@dims <- list("1" = iminfo)
  
  return(se.MSI)
  
}), nm = names(se.MSI.list))

```

As a sanity check, we can plot the MSI coordinates on our new "tissue" image
to see if they align:

```{r fig.width=8, fig.height=2}

plots <- lapply(names(se.MSI.list), function(nm) {
  FeatureOverlay(se.MSI.withIMG.list[[nm]], features = "nFeature_MSI", show.sb = FALSE, pt.size = 1)
})

patchwork::wrap_plots(plots, ncol = 4)
  
```

Now we can merge the two objects `se.RNA` and `se.MSI.filtered` into one single
Seurat object. To do this, we use the `MergeSTData` function

```{r eval=FALSE}

se.merged.list <- lapply(names(se.MSI.withIMG.list), function(nm) {
  se.merged <- MergeSTData(se.RNA.list[[nm]], se.MSI.withIMG.list[[nm]])
  return(se.merged)
})

```

Let's check how that went :-/

Looks like the two coordinate systems are fairly close to each other now.

Let's plot our paired RNA and MSI data points in a spatial map.

```{r fig.width=14, fig.height=2}

plots <- lapply(se.merged.list, function(se.merged) {
  se.merged$type <- ifelse(is.na(se.merged$nCount_MSI), "RNA", "MSI")
  p <- ST.FeaturePlot(se.merged, features = "orig.ident", ncol = 2, cols = "black", 
                      pt.size = 1, label.by = "type")
  return(p)
})

patchwork::wrap_plots(plots, ncol = 4)

```

### Manual alignment
***

With this we should be able to run `ManualAlignImages`. Here we need to 
adjust the MSI section so that it fits with the RNA data section. This 
will be the most time consuming step of this notebook.

To align the data, we need to apply rotation, translations and distortions
using the shiny app opened by `ManualAlignImages`.

NOTE: The tissue outline for samples "V11T17-105_A1" and "V11T17-105_B1" 
are difficult to align, since parts of the tissue are cropped out. The 
alignment of these datasets might therefore be unreliable.

Here are the settings I used:
    For all use "mirror long axis). Up, OK, Up, down
```{r}

transforms <- data.frame(
  sample_id = names(se.MSI.withIMG.list),
  rotation_angle = c(-90, -90,-90, -90),
  move_along_x_axis = c(70, 13, 101, 112 ),
  move_along_y_axis = c(266, 172, 98, 56 ),
  blue_axis_angle = c(0,0,0,0),
  blue_axis_stretch = c(0.96, 0.96, 0.96, 0.96),
  red_axis_angle = c(0,0,0,0),
  red_axis_stretch = c(0.96, 0.96, 0.96, 0.96)
)
transforms

```

Run the alignment function on each dataset.

```{r eval=FALSE}

se.merged.list[[3]] <- ManualAlignImages(se.merged.list[[3]])
se.merged.list[[2]] <- ManualAlignImages(se.merged.list[[2]])
se.merged.list[[4]] <- ManualAlignImages(se.merged.list[[4]])
se.merged.list[[1]] <- ManualAlignImages(se.merged.list[[1]])

```

```{r fig.width=10, fig.height=3}

se.merged.list <- setNames(se.merged.list, nm = names(se.RNA.list))
plots <- lapply(names(se.merged.list), function(nm) {
  se.merged <- se.merged.list[[nm]]
  se.merged$sample_type <- ifelse(GetStaffli(se.merged)@meta.data$sample == "1", "RNA", "MSI")
  p <- ST.FeaturePlot(se.merged, features = "sample_type", ncol = 2, pt.size = 1.5) &
    theme(legend.position = "none", plot.background = element_rect(fill = "lightgray")) &
    labs(title = nm)
  return(p)
})

patchwork::wrap_plots(plots, ncol = 4)

```

Plot them in column to check x dimension: Here it seems that msi data are a bit larger, but I don't want to change the squeeze as I assume that the points are all equidistant

```{r fig.width=10, fig.height=3}

se.merged.list <- setNames(se.merged.list, nm = names(se.RNA.list))
plots <- lapply(names(se.merged.list), function(nm) {
  se.merged <- se.merged.list[[nm]]
  se.merged$sample_type <- ifelse(GetStaffli(se.merged)@meta.data$sample == "1", "RNA", "MSI")
  p <- ST.FeaturePlot(se.merged, features = "sample_type", ncol = 1, pt.size = 1.5) &
    theme(legend.position = "none", plot.background = element_rect(fill = "lightgray")) &
    labs(title = nm)
  return(p)
})

patchwork::wrap_plots(plots, ncol = 4)

```


Now that the images are aligned, we still have one last step left to put 
the MSI and Visium coordinates in the same "space".

From now on, we'll have the coordinates stored as a data.frame instead of 
using the Seurat objects. Below, we calculate a scaling factor to adjust 
the MSI coordinates to fit into the same coordinate system as the RNA data. 

```{r fig.width=24, fig.height=6}

aligned_coordinates <- do.call(rbind, lapply(names(se.merged.list), function(nm) {
  dims <- GetStaffli(se.merged.list[[nm]])@dims
  sf <- dims$`1`$width/dims$`2`$width
  coords <- cbind(se.merged.list[[nm]][[]], GetStaffli(se.merged.list[[nm]])@meta.data) |>
    tibble::rownames_to_column(var = "barcode") |>
    mutate(barcode = gsub(replacement = "_1", x = barcode, pattern = "_\\d")) |>
    select(barcode, warped_x, warped_y, nFeature_MSI) |>
    mutate(sample_type = case_when(is.na(nFeature_MSI) ~ "RNA",
                                   !is.na(nFeature_MSI) ~ "MSI")) |>
    mutate(warped_x = case_when(sample_type == "MSI" ~ warped_x*sf, TRUE ~ warped_x),
           warped_y = case_when(sample_type == "MSI" ~ warped_y*sf, TRUE ~ warped_y),
           sample_id = nm) |>
    select(barcode, warped_x, warped_y, sample_id, sample_type)
  return(coords)
}))

p <- ggplot(aligned_coordinates, aes(warped_x, warped_y,color = sample_type)) +
  geom_point(size = 0.5) +
  facet_wrap(~sample_id, ncol = 4) +
  scale_colour_manual(name="",  
                      values = c("MSI"="orange", "RNA"="blue"))

p

```

```{r eval=FALSE}

# Export plot
p <- ggrastr::rasterize(p, layers = "Point", dpi = 300, dev = "ragg")
pdf("results/plots/aligned_coordinates_spatial_plot.pdf", width = 12, height = 6)
print(p)
dev.off()

```

## data integration
***

Now we just need to come up with a way of "integrating" the data. 
I can think of two simple strategies:

  1. For the MSI data pixels, assign at least 1 nearest neighbor from 
  the RNA data spots and obtain the average expression for each gene
  2. For the MSI data pixels, assign up to k nearest neighbors from the 
  RNA data spots and obtain the average expression for each gene

We also need to apply some kind of distance threshold to remove spots that are too far from a neighbor in the other data set.

First, let's calculate the pairwise distances between the two data sets:

```{r eval=FALSE}

# Split coordinates into a list
aligned_coordinates_list <- split(aligned_coordinates, aligned_coordinates$sample_id)

# Find 5 closest spatial neighbors 
knn_spatial_df_list <- lapply(aligned_coordinates_list, function(coords) {
  
  # get spatial coordinates for RNA and MSI data
  set1 = subset(coords, sample_type == "RNA")[, c("barcode", "warped_x", "warped_y")] |>
    data.frame(row.names = 1)
  set2 = subset(coords, sample_type == "MSI")[, c("barcode", "warped_x", "warped_y")] |>
    data.frame(row.names = 1)
  
  # Calculate pairwise distances using k nearest neighbors algorithm
  knn_spatial_df <- kNN(x = set1, k = 5, query = set2)
  
  # Create a data.frame with MSI pixel IDs (from) pointing to its closest
  # Visium spot (to) and the distance between the two (distance)
  knn_spatial_df <- data.frame(from = rownames(set2)[rep(1:nrow(knn_spatial_df$id), 5)],
                                 to = rownames(set1)[as.vector(knn_spatial_df$id)],
                                 distance = as.vector(knn_spatial_df$dist))
  
  # Add spatial coordinates for "from" and "to"
  knn_spatial_df <- knn_spatial_df |>
    mutate(x = set2[from, "warped_x"],
           x_end = set1[to, "warped_x"],
           y = set2[from, "warped_y"],
           y_end = set1[to, "warped_y"])
  
  return(knn_spatial_df)
})

```

What we see in the plot below are the MSI data points in red 
and their closest RNA data neighbors in blue. We need to get 
rid of the MSI data points where we don't have any RNA data points.

```{r fig.width=12, fig.height=3}

plots <- lapply(names(knn_spatial_df_list), function(nm) {
  knn_spatial_df <- knn_spatial_df_list[[nm]]
  p <- ggplot() +
    geom_segment(data = knn_spatial_df, aes(x = x, xend = x_end, y = y, yend = y_end)) +
    geom_point(data = knn_spatial_df, aes(x, y), color = "red", size = 0.5) +
    geom_point(data = knn_spatial_df, aes(x_end, y_end), color = "blue", size = 0.5) +
    scale_y_reverse() +
    labs(title = nm)
  return(p)
})

p <- patchwork::wrap_plots(plots, ncol = 4)
p

```

```{r eval=FALSE, include=FALSE}

# Export plot
# this plot gets too big with PDF format because of all of the segments
for (i in seq_along(plots)) {
  jpeg(paste0("results/plots/nearest_neighbor_plots/", names(se.MSI.list)[i], ".jpeg"), width = 2000, height = 2000, res = 300)
  print(plots[[i]])
  dev.off()
}

```


If we look at the distance histogram, it should be relatively easy 
to spot what data points are located too far away from each other. 
Somewhere around 35 looks like a reasonable cutoff based on the 
distance histogram below. 

```{r}

gg <- do.call(rbind, lapply(names(knn_spatial_df_list), function(nm) {
  knn_spatial_df <- knn_spatial_df_list[[nm]]
  knn_spatial_df$sample_id <- nm
  return(knn_spatial_df)
}))

ggplot(gg, aes(distance)) +
  geom_histogram(bins = 100) +
  scale_x_log10() +
  geom_vline(xintercept = 35, linetype = "dashed", color = "red")

```

Let's filter out data points that do not have a close match in the other data set:

```{r fig.width=5, fig.height=5}

knn_spatial_df_filtered_list <- lapply(knn_spatial_df_list, function(knn_spatial_df) {
  knn_spatial_df <- knn_spatial_df |>
    dplyr::filter(distance < 35)
  return(knn_spatial_df)
})

```

```{r fig.width=12, fig.height=3}

plots <- lapply(names(knn_spatial_df_filtered_list), function(nm) {
  knn_spatial_df <- knn_spatial_df_filtered_list[[nm]]
  p <- ggplot() +
    geom_segment(data = knn_spatial_df, aes(x = x, xend = x_end, y = y, yend = y_end)) +
    geom_point(data = knn_spatial_df, aes(x, y), color = "red", size = 0.5) +
    geom_point(data = knn_spatial_df, aes(x_end, y_end), color = "blue", size = 0.5) +
    scale_y_reverse() +
    labs(title = nm)
  return(p)
})

p <- patchwork::wrap_plots(plots, ncol = 4)
p

```

```{r eval=FALSE, include=FALSE}

# Export plot
# this plot gets too big with PDF format because of all of the segments
for (i in seq_along(plots)) {
  jpeg(paste0("results/plots/nearest_neighbor_filtered_plots/", names(knn_spatial_df_filtered_list)[i], ".jpeg"), width = 2000, height = 2000, res = 300)
  print(plots[[i]])
  dev.off()
}

```

First we can filter the data to keep only 1 RNA data neighbor per MSI data 
point. This will lead to some loss of data, but is a quite conservative approach.
An alternative would be to average the data between nearest neighbors, thus
interpolating between data points.

```{r}

knn_spatial_df_filtered_list <- setNames(lapply(names(knn_spatial_df_filtered_list), function(nm) {
  
  knn_spatial_df_filtered <- knn_spatial_df_filtered_list[[nm]]
  
  res <- knn_spatial_df_filtered |> 
    dplyr::arrange(from, distance) |>
    dplyr::group_by(from) |>
    dplyr::slice_head(n = 1)
  
  sprintf("Spots removed from sample %s: %s", nm, sum(duplicated(res$to)))
  
  # Remove duplicated Visium spots 
  res <- res[!duplicated(res$to), ]
  
  return(res)
  
}), nm = names(knn_spatial_df_list))

```

Now we can see that every data point in the MSI data has a match in the RNA data. 
However, with this current approach you can see that the regular pattern of the two
data types lead to a loss of data in in certain parts of the tissue. Interpolation
between data points could be used to avoid this.

```{r fig.width=12, fig.height=3}

plots <- lapply(names(knn_spatial_df_filtered_list), function(nm) {
  knn_spatial_df <- knn_spatial_df_filtered_list[[nm]]
  p <- ggplot() +
    geom_segment(data = knn_spatial_df, aes(x = x, xend = x_end, y = y, yend = y_end)) +
    geom_point(data = knn_spatial_df, aes(x, y), color = "red", size = 0.5) +
    geom_point(data = knn_spatial_df, aes(x_end, y_end), color = "blue", size = 0.5) +
    scale_y_reverse() +
    labs(title = nm)
  return(p)
})

p <- patchwork::wrap_plots(plots, ncol = 4)
p

```

```{r eval=FALSE, include=FALSE}

# Export plot
# this plot gets too big with PDF format because of all of the segments
for (i in seq_along(plots)) {
  jpeg(paste0("results/plots/nearest_neighbor_1RNA_1MSI_plots/", names(knn_spatial_df_filtered_list)[i], ".jpeg"), width = 2000, height = 2000, res = 300)
  print(plots[[i]])
  dev.off()
}

```

### Wrapping up alignment process
***

Now that we have our aligned RNA and MSI coordinates, let's export these
as tables:

```{r}

for (nm in names(knn_spatial_df_filtered_list)) {
  write.csv(x = knn_spatial_df_filtered_list[[nm]], file = paste0("results/tables/", nm, "_nearest_neighbors.csv"), quote = FALSE, row.names = FALSE)
}

```

## Create multimodal assay
***

Now we can create a new `Seurat` object where we can analyze 
the two modalities jointly. Here we will simply add the MSI 
data to our RNA Seurat object.

Create a new Seurat object which holds both data modalities.

```{r eval=FALSE}

se.multi.list <- setNames(lapply(names(se.merged.list), function(nm) {
  knn_spatial_df_filtered <- knn_spatial_df_filtered_list[[nm]]
  se.multi <- se.RNA.list[[nm]]
  se.multi <- SubsetSTData(se.multi, spots = knn_spatial_df_filtered$to)
  MSI <- se.MSI.list[[nm]]@assays$MSI@counts[, knn_spatial_df_filtered$from]
  colnames(MSI) <- knn_spatial_df_filtered$to
  se.multi[["MSI"]] <- CreateAssayObject(counts = MSI)
  return(se.multi)
}), nm = names(se.merged.list))

```

Now we can also recalculate some basic QC metrics.

```{r eval=FALSE}

se.multi.list <- lapply(se.multi.list, function(se.multi) {
  se.multi$nFeature_RNA <- colSums(GetAssayData(se.multi, slot = "counts", assay = "RNA") > 0)
  se.multi$nFeature_MSI <- colSums(GetAssayData(se.multi, slot = "counts", assay = "MSI") > 0)
  return(se.multi)
})

```

## Joint analysis
***

NOTE: I selected "CLR" normalization for the MSI data, but another 
normalization method might be more appropriate.

```{r eval=FALSE}

se.multi.list <- lapply(names(se.multi.list), function(nm) {
  
  se.multi <- se.multi.list[[nm]]
  cat(sprintf("Running analysis for sample %s", nm))
  
  # Normalize RNA data
  DefaultAssay(se.multi) <- 'RNA'
  se.multi <- NormalizeData(se.multi, verbose = FALSE) |> 
    FindVariableFeatures(verbose = FALSE) |> 
    ScaleData(verbose = FALSE) |> 
    RunPCA(verbose = FALSE) |>
    RunUMAP(reduction = "pca", dims = 1:30, reduction.name = "RNA.umap", verbose = FALSE) |>
    FindNeighbors(reduction = "pca", dims = 1:30, verbose = FALSE) |>
    FindClusters(verbose = FALSE)
  se.multi$RNA_clusters <- se.multi$seurat_clusters
  
  # Normalize MSI data
  DefaultAssay(se.multi) <- 'MSI'
  se.multi <- NormalizeData(se.multi, normalization.method = "CLR", verbose = FALSE) |> 
    FindVariableFeatures(nfeatures = 200, verbose = FALSE) |> 
    ScaleData(verbose = FALSE) |> 
    RunPCA(reduction.name = "msipca", dims = 1:10, verbose = FALSE) |>
    RunUMAP(reduction = "msipca", dims = 1:10, reduction.name = "MSI.umap", verbose = FALSE) |>
    FindNeighbors(reduction = "msipca", dims = 1:10, verbose = FALSE) |>
    FindClusters(verbose = FALSE)
  se.multi$MSI_clusters <- se.multi$seurat_clusters
  
  return(se.multi)
})

```

Find proper resolution
```{r}

se.multi.list <- lapply(names(se.multi.list), function(nm) {
  
  se.multi <- se.multi.list[[nm]]
  
  DefaultAssay(se.multi) <- 'RNA'
  se.multi <- FindClusters(resolution= 0.01, verbose = T)
  se.multi$RNA_clusters <- se.multi$seurat_clusters
  
  # Normalize MSI data
  DefaultAssay(se.multi) <- 'MSI'
  se.multi <- FindClusters(resolution= 0.01, verbose = T)
  se.multi$MSI_clusters <- se.multi$seurat_clusters
  
  return(se.multi)
})

```


We can look at the UMAPs defined by either RNA or MSI data. It is clear that the RNA
data is more complex it is still possible to find clustered data points in the
MSI data as well.

```{r fig.width=8, fig.height=3}

plots <- lapply(se.multi.list, function(se.multi) {
  p <- DimPlot(se.multi, group.by = "sample_id", reduction = "RNA.umap") | 
    DimPlot(se.multi, group.by = "sample_id", reduction = "MSI.umap")
  return(p)
})

patchwork::wrap_plots(plots, ncol = 1)



```

### Joint clustering
***

see [vignette](https://satijalab.org/seurat/articles/weighted_nearest_neighbor_analysis.html) for more details.

For each cell, we calculate its closest neighbors in the dataset based on a 
weighted combination of RNA and protein similarities. The cell-specific modality 
weights and multimodal neighbors are calculated in a single function, which takes 
~2 minutes to run on this data set. We specify the dimensionality of each modality 
(similar to specifying the number of PCs to include in clustering, but you can 
vary these settings to see that small changes have minimal effect on the overall 
results.

```{r}

# Identify multimodal neighbors. These will be stored in the neighbors slot, 
# and can be accessed using se.multi[['weighted.nn']]
# The WNN graph can be accessed at se.multi[["wknn"]], 
# and the SNN graph used for clustering at se.multi[["wsnn"]]
# Cell-specific modality weights can be accessed at se.multi$RNA.weight

se.multi.list <- lapply(se.multi.list, function(se.multi) {
  se.multi <- FindMultiModalNeighbors(
    se.multi, reduction.list = list("pca", "msipca"), 
    dims.list = list(1:30, 1:10), modality.weight.name = "RNA.weight"
  )
  return(se.multi)
})

```

We can now use these results for downstream analysis, such as visualization and 
clustering. For example, we can create a UMAP visualization of the data based on 
a weighted combination of RNA and protein data We can also perform graph-based 
clustering and visualize these results on the UMAP, alongside a set of cell annotations.

```{r eval=FALSE}

se.multi.list <- lapply(se.multi.list, function(se.multi) {
  se.multi <- RunUMAP(se.multi, nn.name = "weighted.nn", reduction.name = "wnn.umap", reduction.key = "wnnUMAP_", n.epochs = 1e3, min.dist = 0.15)
  se.multi <- FindClusters(se.multi, graph.name = "wsnn", algorithm = 3, resolution = 2, verbose = FALSE)
  se.multi$MM_clusters <- se.multi$seurat_clusters
  return(se.multi)
})

```

Now we can sort the clusters by decreasing RNA weight.

```{r}

plots <- lapply(se.multi.list, function(se.multi) {
  p <- VlnPlot(se.multi, features = "RNA.weight", sort = TRUE)
  return(p)
})

patchwork::wrap_plots(plots[[3]], ncol = 1)

```

And we can compare the results for RNA only, MSI only and RNA+MSI side by side

```{r fig.width=12, fig.height=4}

plots <- lapply(se.multi.list, function(se.multi) {
  p1 <- DimPlot(se.multi, reduction = "RNA.umap", group.by = "RNA_clusters")
  p2 <- DimPlot(se.multi, reduction = "MSI.umap", group.by = "MSI_clusters")
  p3 <- DimPlot(se.multi, reduction = "wnn.umap", group.by = "MM_clusters")
  p <- p1 + p2 + p3 + patchwork::plot_layout(ncol = 3)
  return(p)
})

```

```{r fig.width=12, fig.height=4}

plots <- lapply(se.multi.list, function(se.multi) {
  p1 <- DimPlot(se.multi, reduction = "RNA.umap", group.by = "RNA_clusters")
  p2 <- DimPlot(se.multi, reduction = "MSI.umap", group.by = "MSI_clusters")
  p3 <- DimPlot(se.multi, reduction = "wnn.umap", group.by = "MM_clusters")
  p <- p1 + p2 + p3 + patchwork::plot_layout(ncol = 3)
  return(p)
})

```


```{r fig.width=2, fig.height=2}

DefaultAssay(se.multi.list[[1]])
FeatureOverlay(se.multi.list[[4]], features = "MM_clusters", pt.size = 2)

```

```{r eval=FALSE, include=FALSE}

saveRDS(se.multi.list, file = paste0(working.dir,"/R_objects/hPDStr.multi.list"))

```

# Date
***

```{r}
date()
```

# Session
***

```{r}
devtools::session_info()
```


